\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-09-29 coqimpl.tex
took a pass on coqimpl

However, we later still need the concrete eliminator on "tm" for doing
exhaustiveness checking (as in \ref{chg:extensible-inductive-type}),
thus we will store the corresponding info somewhere.  Here we show one
exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module
"tm_rec_425". This concrete eliminator will be used by our "FRecursor".
For example, in \cref{fig:plugin-example2}, "FRecursor" command will
feed Coq a module "v_33_34" for doing type-checking with specified
handlers (e.g. "neg_handler") during compilation---what it does is
basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.\YZ{
  Sounds like we can still claim the plugin is doing type checking.
  It's just that the checking is done by generating some Coq code and
  then having Coq check it.
}\EDJreply{Yes! If my phrasing doesn't make things clear (i.e. our plugin is doing typechecking),
feel free to change the phrasing for me}
Note that, at this point of the example,
"FRecursor" will only do type check. The compilation of "FRecursor" into
a component of the compiled module only happens when closing the whole
family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and
the (vanilla) inductive type in that compiled module. Thus a "FRecursor"
field is unlike other fields, even though it is exposed as a plain
function for the succeeding fields in the programming interface.\YZ{
  Why is this special treatment needed for compiling FRecursor?
}\EDJreply{because the non-special treatment cannot be applied to recursor --
  the non-special treatment will require the recursor to be *exported* at the interface}

@@

% explain ftheorem wrapping this complicated recursor
This design of decoupling of the case handlers and exhaustiveness checking handles the semantic well, however, it brings a lot of boilerplate code---we need to create an extra (handler) family, manually specifying the type of each case handlers, and then using "FRecursor" to ``tie the knot'' and construct a recursor field from the handlers.  
We thus provide a command "FTheorem" to avoid these boilerplate in programming interface when using tactic programming. 
This command (1) can avoid specifying the handlers family and the type of each case handlers when writing induction like above, (2) invoke
proof interaction mode and thus allow tactic programming, (3) and is also open to extension like "FRecursor".\YZ{example?}\EDJreply{FTheorem is really just a short hand for "FRecursor" + handler family. Nothing more conceptual here. If the reader is interested in the concrete syntax, I think I need to make the two examples avaliable in the appendix.} We expect this "FTheorem" to be used in theorem proving but not general programming.

To summarize, (1) to make each recursor
handler inheritable, we need to seal an abstraction around the inductive
type (e.g., module "b_3"); but to construct a recursor (and carry out the
exhaustiveness checking), we need to break this abstraction and see the
concrete definition of the inductive type (e.g., module "b_rec_12"). Our
meta-theory need to handle these two seemingly contradicting ideas
simultaneously when solving \ref{chg:extensible-inductive-type} . (2) \ref{chg:extensible-inductive-type} is also resolved by this \textit{decoupling} of the syntax of implementation of recursor handlers and exhaustiveness checking. The former is handled by family inheritance and thus avoiding the boilerplate code; and the non-inheritance of the compiled recursor ensures the exhaustiveness checking happens for every recursor.

@@

@@

\textit{Overridable pins Inductive Type and FRecursor}. We notice that, the power of "FRecursor" comes from its distinction from the mundane fields and its ability to stab through the abstraction of "FInductive". Interestingly, the latter is quite similar on how "Overridable .. pins {..}" can see through the abstraction of former pinned overrdiable fields. This inspires us to generalize "pins"---we support pinning inductive type and "FRecursor". Once "FInductive" and "FRecursor" fields are pinned, we are directly exposed with the primitive coq-providing recursor and \textbf{judgemental equality} of the computational information for "FRecursor".

The compilation mechanism is simply make the underlying, compiled definition of inductive type and recursor included as part of the parameters of the module, just like how the concrete definition of overridable fields are included inside the parameters when pinned. This way, the computational information from vanilla Coq can be transimitted into the context.

@@

For example, given a new inductive type, to prove "discriminate" between two constructors, i.e., "c$_i$ .. $\neq$ c$_j$ .." for two different constructors, we simply use the eliminator to map each constructor to different natural number, assuming we can "discriminate" different natural number.\footnote{Rigorously, to avoid any suspicious circular reasoning, we can use large elimination to map them to $\top$ and $\bot$, and because $\exists a : \_, \top$ holds for the former not the later, thus $\top \neq \bot$.} "injection" can be supported in a similar way---we can simply let eliminator map each inductive type to the arguments of the constructors.

@@

Thus we alter eliminators into \text{partial recursor}, such that (1) it will not hinder the future extension, and (2) it is still powerful enough to derive injectiveness like tactic "inject" and "discriminate". What's more, (3) we can actually prove this \text{partial recursor} is actually good enough because it can act as an \textit{extensional characterization} for extensible inductive type \textit{to some extent}. We will call it \textit{propositional} because we only prpositionally postulate its computational informations for Coq.

@@

% This raise an interesting question: what propositions are conservative extension? 
% For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a \textbf{``best'' proposition} (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

% Since injectiveness of the constructors are derived by recursor in vanilla inductive type, an educated guess would be related to recursor. But we know the original recursor won't work because the mere existence of the original recursor will break extensibility. Thus we need to modify it.
% The simplest idea is to allow partiality---i.e., returning
% \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

% We call this \textit{propositional partial recursor}.\YZ{What is propositional about partial recursors? I suppose it has to do with the computational axioms?}\EDJreply{Yes. I want to emphasize the equality used by computational axiom is not judgemental equality. So Coq cannot automatically compute/reduce when a partial recursor applied to a term. This reduction can only be semi-automatically done by using [rewrite tactic and the computational axiom].}
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is\YZ{Mention that partial recursors can be automatically generated}\EDJreply{Added below please check.}

\begin{minted}[escapeinside=@@]{Coq}
(* Inside Family B *)
FInductive b : Set := tt : b | ff : b.
b_prec : forall R, option R -> option R -> b -> option R.@\YZ{Not clear what it takes to define/prove partial recursors. The reader needs to be told if they require breaking abstraction like recursors do.}\EDJreply{What do you mean? recursor doesn't break abstraction but break extensibility. Below mentioned partial recursor doesn't break extensibility}@
\end{minted}
with appropriate computational axioms. This partial recursor differs from original recursor only by decorating an "option" with the return type thus possible to be automatically generated (and the case is similar for those computational axiom). Thanks to the partiality, partial recursor doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec") because for example, "b_prec" can just return "None" when bump into extended constructor in the future.

The second thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{B.tt ≠
B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : B.tt = B.ff → False.
(* Because B.tt = B.ff → reflect B.tt = reflect B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

Thirdly, even though we cannot prove partial recursor is \textit{the best} reasoning tool for extensible inductive type,
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (under certain constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), .., 
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of
  "C" and "prec" of "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
The reason we have this weird looking \ref{prop:detectable-partiality} is because the construction of "linv"---at one point, using "prec" to construct "linv", we need to prove \mintinline{Coq}{∀ a, (Bᵢ a → option C) → option C}, generally not provable, only if we have \mintinline{Coq}{option (Bᵢ a → C)} as an argument instead. This \ref{prop:detectable-partiality} is exactly transforming former to the later: recall that, according to the semantic of inductive type, \mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using "linv" on the substructure of "D". This function points out each of the translation result from "D" to "C" and some of them are "None" because of failing. For us, we only care about if \textbf{any} of the substructure of "D" is failed to translate---if any one substructure fails, then the whole transltion fails and thus we should have "None" as return.

>>>>

However, we later still need the concrete eliminator on "tm" for doing
exhaustiveness checking (as in \ref{chg:extensible-inductive-type}),
thus we will store the corresponding info somewhere.  Here we show one
exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module
"tm_rec_425". This concrete eliminator will be used by our "FRecursor".
For example, in \cref{fig:plugin-example2}, "FRecursor" command will
feed Coq a module "v_33_34" for doing type-checking with specified
handlers (e.g. "neg_handler") during compilation---what it does is
basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.\YZ{
  Sounds like we can still claim the plugin is doing type checking.
  It's just that the checking is done by generating some Coq code and
  then having Coq check it.
}\EDJreply{Yes! If my phrasing doesn't make things clear (i.e. our plugin is doing typechecking),
feel free to change the phrasing for me}
Note that, at this point of the example,
"FRecursor" will only do type check. The compilation of "FRecursor" into
a component of the compiled module only happens when closing the whole
family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and
the (vanilla) inductive type in that compiled module. Thus a "FRecursor"
field is unlike other fields, even though it is exposed as a plain
function for the succeeding fields in the programming interface.

@@

% explain ftheorem wrapping this complicated recursor
This design of decoupling of the case handlers and exhaustiveness
checking handles the semantic well, however, it brings a lot of
boilerplate code---we need to create an extra (handler) family, manually
specifying the type of each case handlers, and then using "FRecursor" to
``tie the knot'' and construct a recursor field from the handlers.  
We thus provide a command "FTheorem" to avoid these boilerplate in programming interface when using tactic programming. 
This command (1) can avoid specifying the handlers family and the type of each case handlers when writing induction like above, (2) invoke
proof interaction mode and thus allow tactic programming, (3) and is also open to extension like "FRecursor". We expect this "FTheorem" to be used in theorem proving but not general programming.

To summarize, (1) to make each recursor
handler inheritable, we need to seal an abstraction around the inductive
type (e.g., module "b_3"); but to construct a recursor (and carry out the
exhaustiveness checking), we need to break this abstraction and see the
concrete definition of the inductive type (e.g., module "b_rec_12"). Our
meta-theory need to handle these two seemingly contradicting ideas
simultaneously when solving \ref{chg:extensible-inductive-type}.
(2) \ref{chg:extensible-inductive-type} is also resolved by this
\textit{decoupling} of the syntax of implementation of recursor handlers
and exhaustiveness checking. The former is handled by family inheritance
and thus avoiding the boilerplate code; and the non-inheritance of the
compiled recursor ensures the exhaustiveness checking happens for every
recursor.

@@

\paragraph{Tactics and propositional recursors.}

@@

\textit{Overridable pins Inductive Type and FRecursor}. We notice that,
the power of "FRecursor" comes from its distinction from the mundane
fields and its ability to stab through the abstraction of "FInductive".
Interestingly, the latter is quite similar to how "Overridable .. pins
{..}" can see through the abstraction of former pinned overrdiable
fields. This inspires us to generalize "pins"---we support pinning
inductive type and "FRecursor". Once "FInductive" and "FRecursor" fields
are pinned, we are directly exposed with the primitive coq-providing
recursor and \textbf{judgemental equality} of the computational
information for "FRecursor".

The compilation mechanism is to simply make the underlying, compiled
definition of inductive type and recursor included as part of the
parameters of the module, just like how the concrete definition of
overridable fields are included inside the parameters when pinned. This
way, the computational information from vanilla Coq can be transmitted
into the context.

@@

For example, given a new inductive type, to prove "discriminate" between two constructors, i.e., "c$_i$ .. $\neq$ c$_j$ .." for two different constructors, we simply use the eliminator to map each constructor to a different natural number, assuming we can "discriminate" different natural numbers.\footnote{Rigorously, to avoid any suspicious circular reasoning, we can use large elimination to map them to $\top$ and $\bot$, and because $\exists a : \_, \top$ holds for the former not the latter, thus $\top \neq \bot$.} "injection" can be supported in a similar way---we can simply let eliminator map each inductive type to the arguments of the constructors.

@@

Thus we alter eliminators into \text{partial recursor}, such that (1) it
will not hinder the future extension, and (2) it is still powerful
enough to derive injectivity ("inject") and disjointness ("discriminate") of constructors.
What's more, (3) we can prove this \text{partial recursor} is
actually good enough because it can act as an \textit{extensional
characterization} for extensible inductive types \textit{to some extent}.
We will call it \textit{propositional} because we only propositionally
postulate its computational information for Coq.

@@

% This raise an interesting question: what propositions are conservative extension? 
% For example, apparently we know injectivity of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a \textbf{``best'' proposition} (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

% Since injectivity of the constructors are derived by recursor in vanilla inductive type, an educated guess would be related to recursor. But we know the original recursor won't work because the mere existence of the original recursor will break extensibility. Thus we need to modify it.
% The simplest idea is to allow partiality---i.e., returning
% \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

% We call this \textit{propositional partial recursor}.\YZ{What is propositional about partial recursors? I suppose it has to do with the computational axioms?}\EDJreply{Yes. I want to emphasize the equality used by computational axiom is not judgemental equality. So Coq cannot automatically compute/reduce when a partial recursor applied to a term. This reduction can only be semi-automatically done by using [rewrite tactic and the computational axiom].}
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is

\begin{minted}[escapeinside=@@]{Coq}
(* Inside Family B *)
FInductive b : Set := tt : b | ff : b.
b_prec : forall R, option R -> option R -> b -> option R.
\end{minted}
with appropriate computational axioms.  This partial
recursor differs from original recursor only by decorating the return
type with an "option" and can thus be automatically generated (and
the case is similar for those computational axiom). Thanks to the
partiality, partial recursor doesn't break
extensibility (i.e., all future extensions of "b" can support this
"b_prec") because for example, "b_prec" can just return "None" when bump into extended constructor in the future.

The second thing to check is that, if partial recursor is theoretically enough for injectivity of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{B.tt ≠
B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : B.tt = B.ff → False.
(* Because B.tt = B.ff → reflect B.tt = reflect B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

Thirdly, even though we cannot prove partial recursors are \textit{the
best} reasoning tool for extensible inductive types,
the above "reflect" function implies that propositional partial
recursors are a good \textit{extensional characterization} of
(non-indexed) extensible inductive type (under certain
constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
The reason we have this weird looking \ref{prop:detectable-partiality}
is that the construction of "linv"---at one point, using "prec" to
construct "linv", we need to prove \mintinline{Coq}{∀ a, (Bᵢ a → option
C) → option C}, generally not provable, unless we have
\mintinline{Coq}{option (Bᵢ a → C)} as an argument instead.\YZ{Don't you really mean
"unless we have the Detectable Partiality assumption instead"?}
This \ref{prop:detectable-partiality} is exactly transforming former to
the latter:
recall that, according to the semantic of inductive type,
\mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using
"linv" on the substructure of "D". This function points out each of the
translation result from "D" to "C" and some of them are "None" because
of failing. For us, we only care about if \textbf{any} of the
substructure of "D" fails to translate---if any substructure fails,
then the whole translation fails and thus we should have "None" as
return.
